Artificial ecosystem selection reveals relationships between microbiome composition and ecosystem function

### 2023-02-27 manuscript/manuscript.Rmd
clean up references and update figure captions

{title max characters 150; current = 106 w/spaces}

^1^ Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA  
\* Correspondence: Andrew H. Morris, amorris3\@uoregon.edu  
Competing Interests: We declare we have no competing interests.

@@

load(file = paste0(der_dir, 'plots.RData'))
load(file = paste0(der_dir, 'models.RData'))

@@

{max 250 words; current = 249}

Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know _a priori_ which
microbiome members influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil methane oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
methane oxidation rate per passage. We
estimated that 40\% of the variation in methane oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known methane oxidizers; instead, `r length(da_fams)` families not
known to oxidize methane, including *Fimbriimonadaceae*, *Cytophagaceae*, and
*Diplorickettsiaceae*, were enriched by selection. This result is in contrast to the typical assumption that
the rate of an ecosystem function is limited by the final step in the associated
microbial pathway. Our study demonstrates that variation in microbiome
composition can contribute to variation in the rate of ecosystem function and
that manipulating microbiome composition could be a viable strategy for managing
ecosystem functions.

@@

This relationship between microbiome variation and ecosystem function variation
is analogous to the concept of "heritability" [@lynch1998] used by quantitative
geneticists, or more precisely the concept of "microbiability" [@difford2016] proposed
by microbiome scientists who study host-associated microbiomes. Although rarely
used in the study of environmental microbiomes, this concept could be very
useful for understanding and manipulating microbially-mediated functions in a
variety of ecosystems. In our experiment, variation in microbiome taxonomic
composition statistically explained (i.e., was associated with) 40% of the
variation we observed in the rate of methane oxidation.  This suggests that
there is substantial potential for altering this ecosystem function through
microbiome manipulation in this soil. It is very likely that the
“environmental microbiability” will be different for other ecosystem functions
in this soil and for methane oxidation in other soils. However, our experiment
demonstrates that this relationship is measurable and provides an example of how
this can be accomplished.

@@

# Figures

![](Figure1.png)

@@

```{r response, fig.width=4, fig.height=3}

p_response

`r figures("response", "Response to selection on soil CH~4~ oxidation rate. The y-axis is the log~10~ of the additive inverse of the first-order exponential decay constant $k$ (i.e., $-k$) with units day^-1^ so that a more positive value represents a higher CH~4~ oxidation rate.")`

```{r herit, fig.width=7}
# 
herit_plot
#`r figures("herit", paste0("Ecosystem heritability as the regression of mid-recipient methane flux on mid-donor methane flux ($-k$ with units day^-1^). Donor is the mean oxidation rate of the three jars selected to inoculate the next passage. Recipient is the mean oxidation rate of the 12 jars in the next passage. The orange lines and points are the Positive treatment and the gray lines and points are the Neutral treatment. Slope of the regression line estimates the proportion of variation in methane oxidation rate due to variation in the microbiome. For the Positive treatment, the variance explained by the microbiome was ", myround((coefficients(herit_model)[2] + coefficients(herit_model)[4])*100, 1), "% (slope = ", myround(coefficients(herit_model)[2] + coefficients(herit_model)[4], 3), ", SE = ", myround(tidy(herit_model)$std.error[4], 2), ") and for the Neutral treatment the variance explained by the microbiome was ", myround(coefficients(herit_model)[2]*100, 1), "% (slope = ", myround(coefficients(herit_model)[2], 3), ", SE = ", myround(tidy(herit_model)$std.error[2], 2), ")."))`
```

`r figures("herit", paste0("Divergence between the positive selection line and the control line over five passages. Values are CH~4~ oxidation rate as the additive inverse of the first-order exponential decay constant k (day^-1^). Regression of divergence on cumulative selection differential provides an estimate of realized heritability (h^2^ &pm; SE) and the slope of the regression was ", myround(tidy(rh_model)[[2, 2]], 2), " &pm; ", myround(tidy(rh_model)[[2, 3]], 2), "."))`

```{r beta, results=FALSE, fig.height=2.5}

p_beta

`r figures("beta", "Non-metric multidimensional scaling plot of beta diversity for all jars. Dissimilarities are based on rarefied Bray-Curtis dissimilarity. Colors represent treatment and shapes represent passage number.")`

```{r da, results=FALSE}

p_log10

`r figures("da", "Differentially abundant family-level taxa identified by ANCOM-II, ALDEx2, and CORNCOB. Values on the x-axis are relative abundances on a log~10~ scale. Large points with error bars are the median relative abundance and interquartile range. Taxa in the top panel are enriched in the Positive treatment and taxa in the bottom panel are depleted in the Positive treatment. Taxa are sorted by their effect size with taxa at the top having the largest positive effect size and taxa at the bottom with the largest negative effect size.")`

>>>>

{title max characters 150; current = 106 w/spaces}  
^1^ Institute of Ecology and Evolution, University of Oregon, Eugene, OR, USA  
\* Correspondence: Andrew H. Morris, amorris3\@uoregon.edu  
Competing Interests: We declare we have no competing interests.

@@

#load(file = paste0(der_dir, 'plots.RData'))
load(file = paste0(der_dir, 'models.RData'))

@@

{max 250 words; current = 249}  
Microbiomes mediate important ecosystem functions, yet it has proven difficult
to determine the relationship between microbiome composition and the rate of
ecosystem functions. This challenge remains because it is difficult to
manipulate microbiome composition directly, we often cannot know _a priori_ which
microbiome members influence the rate of an ecosystem function, and
microbiomes can covary strongly with other drivers of ecosystem function, such
as the environment. We propose that artificial ecosystem selection can overcome
these challenges. Artificial ecosystem selection involves creating replicate
ecosystems, selecting replicates with a desired ecosystem-level trait, and
using the microbiomes from the selected ecosystems to inoculate a new set of
ecosystems. We performed artificial ecosystem selection on replicate soil
microbiomes to select for increased soil methane oxidation rate. We observed a
strong response to selection with a `r myround(per_slope, 1)`\% increase in
methane oxidation rate per passage. We
estimated that `r myround(tidy(rh_model)[[2, 2]]*100, 1)`\% of the variation in
methane oxidation
rate in these soils can be attributed to microbiome variation. We also found
that selection did not enrich for known methane oxidizers; instead, `r length(da_fams)` families not
known to oxidize methane, including *Fimbriimonadaceae*, *Cytophagaceae*, and
*Diplorickettsiaceae*, were enriched by selection. This result is in contrast to the typical assumption that
the rate of an ecosystem function is limited by the final step in the associated
microbial pathway. Our study demonstrates that variation in microbiome
composition can contribute to variation in the rate of ecosystem function and
that manipulating microbiome composition could be a viable strategy for managing
ecosystem functions.

@@

This relationship between microbiome variation and ecosystem function variation
is analogous to the concept of "heritability" [@lynch1998] used by quantitative
geneticists, or more precisely the concept of "microbiability" [@difford2016] proposed
by microbiome scientists who study host-associated microbiomes. Although rarely
used in the study of environmental microbiomes, this concept could be very
useful for understanding and manipulating microbially-mediated functions in a
variety of ecosystems. In our experiment, variation in microbiome taxonomic
composition statistically explained (i.e., was associated with) 
`r myround(tidy(rh_model)[[2, 2]]*100, 1)`\% of the
variation we observed in the rate of methane oxidation.  This suggests that
there is substantial potential for altering this ecosystem function through
microbiome manipulation in this soil. It is very likely that the
“environmental microbiability” will be different for other ecosystem functions
in this soil and for methane oxidation in other soils. However, our experiment
demonstrates that this relationship is measurable and provides an example of how
this can be accomplished.

@@

# Figure Captions

@@

`r figures("response", paste0("Response to selection on soil CH~4~ oxidation rate. The y-axis is CH~4~ oxidation rate as the log~10~ of the additive inverse of the first-order exponential decay constant $k$ (i.e., $-k$) with units day^-1^ so that a more positive value represents a higher CH~4~ oxidation rate. There was a significant difference of slopes between the positive selection treatment and the control (F~", compare_response_mods[2,4], ",", compare_response_mods[2,2], "~ = ", myround(compare_response_mods[2,5], 2), ", p = ", myround(compare_response_mods[2,6], 2), ")."))`

`r figures("herit", paste0("Divergence between the positive selection line and the control line over the cumulative selection differential. Values are CH~4~ oxidation rate as the additive inverse of the first-order exponential decay constant k (day^-1^). Regression of divergence on cumulative selection differential provides an estimate of realized heritability (h^2^ &pm; SE) and the slope of the regression was ", myround(tidy(rh_model)[[2, 2]], 2), " &pm; ", myround(tidy(rh_model)[[2, 3]], 2), ", though this was not significant (F~", glance(rh_model)$df,  ",", glance(rh_model)$df.residual,  "~ = ", myround(glance(rh_model)$statistic, 2),  ", p =  ", myround(glance(rh_model)$p.value, 2),  ")."))`

`r figures("beta", "Non-metric multidimensional scaling plot of beta diversity for all jars. Dissimilarities are rarefied Bray-Curtis dissimilarity averaged over 99 subsamples.")`

`r figures("da", "Differentially abundant family-level taxa identified by ANCOM-II, ALDEx2, and CORNCOB. Values on the x-axis are relative abundances on a log~10~ scale. Taxa in the top panel are enriched in the positive selection treatment relative to the control and taxa in the bottom panel are depleted in the positive selection treatment relative to the control. Taxa are sorted by their effect size with taxa at the top having the largest positive effect size and taxa at the bottom with the largest negative effect size.")`
